Do we need detailed demographic data to forecast plant population responses to climate change?

### 2015-06-09 manuscript/manuscript.Rmd
updates ms with IPM description

Do we need detailed demographic data to forecast population responses to climate change?
========================================================================================

### Andrew T. Tredennick and Peter B. Adler

*Andrew T. Tredennick[^corrauth], Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*

@@

@@

All of our analyses (model fitting and simulating) were conducted in R [@R2013]. We used the 'No-U-Turn' MCMC sampler in Stan [@stan2014]  to estimate the posterior distributions of model parameters using the package 'rstan' [@rstan2014]. We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 1,000 iterations after discarding an initial 1,000 iterations. We recignize such short MCMC chains may surprise those more familiar with other MCMC samplers (i.e. JAGS or WinBUGS), but the Stan sampler is exceptionally efficient, which reduces the number of iterations needed to achieve convergence. We assessed convergence visually and made sure scale reduction factors for all parameters were less than 1.01. For the purposes of including parameter uncertaintyy in our population models, we saved the final 1,000 iterations from each of the three MCMC chains for all parameters to be used as randomly drawn values during population simulation. This step alleviates the need to reduce model parameters by model selection since sampling from the full parameter space in the MCMC ensures that if a parameter broadly overlaps zero, on average the effect in the population models will also be near zero.

### Population models
With the posterior distribution of the vital rate statistical models in hand, it is straightforward to simulate the population models. We used an Integral Projection Model (IPM) to model populations based on individual level data and an quadrat based version of an individually-based model (Quadrat-Based Model, QBM) to model populations based on quadrat level data. Both models take the general form:

\begin{equation}
N_{t+1} = S \times G + R.
\end{equation}

So, at each time step in a simulation, we use the survival regression to determine if each genet lives or not (if each quadrat remains occupied or not), the growth regression to determine size changes of surviving individuals (cover change of occupied quadrats), and the recruitment regression to determine the number of new recruits (if a quadrat is colonized or not). We first use one-step-ahead forecasts to assess each model's ability to reproduce observed cover changes. Then we use the models to analyze the effect of potential climate changes on population size over long time scales.

We used random draws from the final 1,000 iterations from each of three MCMC chains to introduce stochasticity into our population models. At each time step, we randomly selected climate covariates from one of the 14 observed years. Then, we drew the full parameter set (climate effects and density-dependence fixed effects) from a randomly selected MCMC iteration. Using this approach, rather than simply using coefficient point estimates, ensures that relatively unimportant climate covariates (those that broadly overlap 0) have little effect on the simulation results. Since our focus was on the contribution of climate covariates to population states, we set the random year effects and the random group effects to zero.

>>>>

Do we need detailed demographic data to forecast population responses to climate change? Probably.
========================================================================================

### Andrew T. Tredennick[^corrauth] and Peter B. Adler

*Andrew T. Tredennick, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*

@@

All R code and data necessary to replicate our analysis is archived on GitHub as release 1.0 (http://github.com/atredennick/MicroMesoForecast/releases). That stable release will remain static as a record of this analysis, but subsequent versions may appear if we update this work.

@@

All of our analyses (model fitting and simulating) were conducted in R [@R2013]. We used the 'No-U-Turn' MCMC sampler in Stan [@stan2014]  to estimate the posterior distributions of model parameters using the package 'rstan' [@rstan2014]. We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 1,000 iterations after discarding an initial 1,000 iterations. We recignize such short MCMC chains may surprise those more familiar with other MCMC samplers (i.e. JAGS or WinBUGS), but the Stan sampler is exceptionally efficient, which reduces the number of iterations needed to achieve convergence. We assessed convergence visually and made sure scale reduction factors for all parameters were less than 1.01. For the purposes of including parameter uncertainty in our population models, we saved the final 1,000 iterations from each of the three MCMC chains for all parameters to be used as randomly drawn values during population simulation. This step alleviates the need to reduce model parameters by model selection since sampling from the full parameter space in the MCMC ensures that if a parameter broadly overlaps zero, on average the effect in the population models will also be near zero.

### Population models
With the posterior distribution of the vital rate statistical models in hand, it is straightforward to simulate the population models. We used an Integral Projection Model (IPM) to model populations based on individual level data and an quadrat based version of an individually-based model (Quadrat-Based Model, QBM) to model populations based on quadrat level data. We describe each in turn.

#### Integral projection model
We use an environmentally stochastic IPM [@Rees2009] that includes the random year effects and the climate covariates from the vital rate statistical models. But note that we can, and do for some simulations, ignore the random year effects so that only the climate effects can drive interannual variation. Our IPM follows the specification of @Chu2015 where the population of species _j_ is a density function $n(u_{j},t)$ giving the density of sized-_u_ genets at time _t_. Genet size is on the natural log scale, so that $n(u_{j},t)du$ is the number of genets whose area (on the arithmetic scale) is between $e^{u_{j}}$ and $e^{u_{j}+du}$. So, the density function for any size _v_ at time $t+1$ is

\begin{equation}
n(v_{j},t+1) = \int_{L_{j}}^{U_{j}} k_{j}(v_{j},u_{j},\bar{\bold{w_{j}}}(u_{j}))n(u_{j},t)
\end{equation}

where $k_{j}(v_{j},u_{j},\bar{\bold{w_{j}}})$ is the population kernal that describes all possible transitions from size $u$ to $v$ and $\bar{\bold{w_{j}}}$ is a vector of estimates of average crowding experienced from all other species by a genet of size $u_j$ and species $j$. The integral is evaluated over all possible sizes between predefined lower (_L_) and upper (_U_) size limits that extend beyond the range of observed genet sizes.

The population kernal is defined as the joint contributions of survival (_S_), growth (_G_), and recruitment (_R_):

\begin{equation}
k_{j}(v_{j},u_{j},\bar{\bold{w_{j}}}) = S_j(u_j, \bar{\bold{w_{j}}}(u_{j}))G_j(v_{j},u_{j},\bar{\bold{w_{j}}}(u_{j})) + R_j(v_{j},u_{j},\bar{\bold{w_{j}}}),
\end{equation}

which, said plainly, means we are calculating growth (_G_) for individuals that survive (_S_) from time _t_ to _t+1_ and adding in newly recruited (_R_) individuals of an average sized one-year-old genet for the focal species. Our stastical model for recruitment (_R_, described above) returns the number of new recruit produced per quadrat. Following previous work [@Adler2012; @Chu2015], we assume that fecundity increases linearly with size ($R_j(v_{j},u_{j},\bar{\bold{w_{j}}}) = e^{u_j}R_j(v_{j},\bar{\bold{w_{j}}})$) to incorporate the recruitment function in the spatially-implicit IPM.

We used random draws from the final 1,000 iterations from each of three MCMC chains to introduce stochasticity into our population models. We thinned these chains by 10 to reduce potential autocorrelation, leaving us with 300 samples for each parameter. At each time step, we randomly selected climate covariates from one of the 14 observed years. Then, we drew the full parameter set (climate effects and density-dependence fixed effects) from a randomly selected MCMC iteration. Using this approach, rather than simply using coefficient point estimates, ensures that relatively unimportant climate covariates (those that broadly overlap 0) have little effect on the simulation results. Since our focus was on the contribution of climate covariates to population states, we set the random year effects and the random group effects to zero.

#### Quad-based model
Our quad-based model (QBM) perfectly mirrors its statistical description (Eq. #). We use the same approach for drawing parameter values as described for the IPM.

### 2015-06-09 manuscript/manuscript.tex
updates ms with IPM description

\section{Do we need detailed demographic data to forecast population
responses to climate
change?}\label{do-we-need-detailed-demographic-data-to-forecast-population-responses-to-climate-change}

\subsubsection{Andrew T. Tredennick and Peter B.
Adler}\label{andrew-t.-tredennick-and-peter-b.-adler}

\emph{Andrew T. Tredennick\footnote{Corresponding author:
  \href{mailto:atredenn@gmail.com}{\href{mailto:atredenn@gmail.com}{atredenn@gmail.com}}},
Department of Wildland Resources and the Ecology Center, Utah State
University, Logan, UT}

@@

@@

All of our analyses (model fitting and simulating) were conducted in R
(R Core Development Team 2013). We used the `No-U-Turn' MCMC sampler in
Stan (Stan Development Team 2014a) to estimate the posterior
distributions of model parameters using the package `rstan' (Stan
Development Team 2014b). We obtained posterior distributions for all
model parameters from three parallel MCMC chains run for 1,000
iterations after discarding an initial 1,000 iterations. We recignize
such short MCMC chains may surprise those more familiar with other MCMC
samplers (i.e.~JAGS or WinBUGS), but the Stan sampler is exceptionally
efficient, which reduces the number of iterations needed to achieve
convergence. We assessed convergence visually and made sure scale
reduction factors for all parameters were less than 1.01. For the
purposes of including parameter uncertaintyy in our population models,
we saved the final 1,000 iterations from each of the three MCMC chains
for all parameters to be used as randomly drawn values during population
simulation. This step alleviates the need to reduce model parameters by
model selection since sampling from the full parameter space in the MCMC
ensures that if a parameter broadly overlaps zero, on average the effect
in the population models will also be near zero.

@@

With the posterior distribution of the vital rate statistical models in
hand, it is straightforward to simulate the population models. We used
an Integral Projection Model (IPM) to model populations based on
individual level data and an quadrat based version of an
individually-based model (Quadrat-Based Model, QBM) to model populations
based on quadrat level data. Both models take the general form:

\begin{equation}
N_{t+1} = S \times G + R.
\end{equation}

So, at each time step in a simulation, we use the survival regression to
determine if each genet lives or not (if each quadrat remains occupied
or not), the growth regression to determine size changes of surviving
individuals (cover change of occupied quadrats), and the recruitment
regression to determine the number of new recruits (if a quadrat is
colonized or not). We first use one-step-ahead forecasts to assess each
model's ability to reproduce observed cover changes. Then we use the
models to analyze the effect of potential climate changes on population
size over long time scales.

We used random draws from the final 1,000 iterations from each of three
MCMC chains to introduce stochasticity into our population models. At
each time step, we randomly selected climate covariates from one of the
14 observed years. Then, we drew the full parameter set (climate effects
and density-dependence fixed effects) from a randomly selected MCMC
iteration. Using this approach, rather than simply using coefficient
point estimates, ensures that relatively unimportant climate covariates
(those that broadly overlap 0) have little effect on the simulation
results. Since our focus was on the contribution of climate covariates
to population states, we set the random year effects and the random
group effects to zero.

@@

@@

@@

>>>>

\section{Do we need detailed demographic data to forecast population
responses to climate change?
Probably.}\label{do-we-need-detailed-demographic-data-to-forecast-population-responses-to-climate-change-probably.}

\subsubsection[Andrew T. Tredennick and Peter B. Adler]{Andrew T.
Tredennick\footnote{Corresponding author:
  \href{mailto:atredenn@gmail.com}{\href{mailto:atredenn@gmail.com}{atredenn@gmail.com}}}
and Peter B.
Adler}\label{andrew-t.-tredennickcorrauth-and-peter-b.-adler}

\emph{Andrew T. Tredennick, Department of Wildland Resources and the
Ecology Center, Utah State University, Logan, UT}

@@

All R code and data necessary to replicate our analysis is archived on
GitHub as release 1.0
(\url{http://github.com/atredennick/MicroMesoForecast/releases}). That
stable release will remain static as a record of this analysis, but
subsequent versions may appear if we update this work.

@@

All of our analyses (model fitting and simulating) were conducted in R
(R Core Development Team 2013). We used the `No-U-Turn' MCMC sampler in
Stan (Stan Development Team 2014a) to estimate the posterior
distributions of model parameters using the package `rstan' (Stan
Development Team 2014b). We obtained posterior distributions for all
model parameters from three parallel MCMC chains run for 1,000
iterations after discarding an initial 1,000 iterations. We recignize
such short MCMC chains may surprise those more familiar with other MCMC
samplers (i.e.~JAGS or WinBUGS), but the Stan sampler is exceptionally
efficient, which reduces the number of iterations needed to achieve
convergence. We assessed convergence visually and made sure scale
reduction factors for all parameters were less than 1.01. For the
purposes of including parameter uncertainty in our population models, we
saved the final 1,000 iterations from each of the three MCMC chains for
all parameters to be used as randomly drawn values during population
simulation. This step alleviates the need to reduce model parameters by
model selection since sampling from the full parameter space in the MCMC
ensures that if a parameter broadly overlaps zero, on average the effect
in the population models will also be near zero.

@@

With the posterior distribution of the vital rate statistical models in
hand, it is straightforward to simulate the population models. We used
an Integral Projection Model (IPM) to model populations based on
individual level data and an quadrat based version of an
individually-based model (Quadrat-Based Model, QBM) to model populations
based on quadrat level data. We describe each in turn.

\paragraph{Integral projection model}\label{integral-projection-model}

We use an environmentally stochastic IPM (Rees and Ellner 2009) that
includes the random year effects and the climate covariates from the
vital rate statistical models. But note that we can, and do for some
simulations, ignore the random year effects so that only the climate
effects can drive interannual variation. Our IPM follows the
specification of Chu and Adler (2015) where the population of species
\emph{j} is a density function $n(u_{j},t)$ giving the density of
sized-\emph{u} genets at time \emph{t}. Genet size is on the natural log
scale, so that $n(u_{j},t)du$ is the number of genets whose area (on the
arithmetic scale) is between $e^{u_{j}}$ and $e^{u_{j}+du}$. So, the
density function for any size \emph{v} at time $t+1$ is

\begin{equation}
n(v_{j},t+1) = \int_{L_{j}}^{U_{j}} k_{j}(v_{j},u_{j},\bar{\bold{w_{j}}}(u_{j}))n(u_{j},t)
\end{equation}

where $k_{j}(v_{j},u_{j},\bar{\bold{w_{j}}})$ is the population kernal
that describes all possible transitions from size $u$ to $v$ and
$\bar{\bold{w_{j}}}$ is a vector of estimates of average crowding
experienced from all other species by a genet of size $u_j$ and species
$j$. The integral is evaluated over all possible sizes between
predefined lower (\emph{L}) and upper (\emph{U}) size limits that extend
beyond the range of observed genet sizes.

The population kernal is defined as the joint contributions of survival
(\emph{S}), growth (\emph{G}), and recruitment (\emph{R}):

\begin{equation}
k_{j}(v_{j},u_{j},\bar{\bold{w_{j}}}) = S_j(u_j, \bar{\bold{w_{j}}}(u_{j}))G_j(v_{j},u_{j},\bar{\bold{w_{j}}}(u_{j})) + R_j(v_{j},u_{j},\bar{\bold{w_{j}}}),
\end{equation}

which, said plainly, means we are calculating growth (\emph{G}) for
individuals that survive (\emph{S}) from time \emph{t} to \emph{t+1} and
adding in newly recruited (\emph{R}) individuals of an average sized
one-year-old genet for the focal species. Our stastical model for
recruitment (\emph{R}, described above) returns the number of new
recruit produced per quadrat. Following previous work (Adler et al.
2012, Chu and Adler 2015), we assume that fecundity increases linearly
with size
($R_j(v_{j},u_{j},\bar{\bold{w_{j}}}) = e^{u_j}R_j(v_{j},\bar{\bold{w_{j}}})$)
to incorporate the recruitment function in the spatially-implicit IPM.

We used random draws from the final 1,000 iterations from each of three
MCMC chains to introduce stochasticity into our population models. We
thinned these chains by 10 to reduce potential autocorrelation, leaving
us with 300 samples for each parameter. At each time step, we randomly
selected climate covariates from one of the 14 observed years. Then, we
drew the full parameter set (climate effects and density-dependence
fixed effects) from a randomly selected MCMC iteration. Using this
approach, rather than simply using coefficient point estimates, ensures
that relatively unimportant climate covariates (those that broadly
overlap 0) have little effect on the simulation results. Since our focus
was on the contribution of climate covariates to population states, we
set the random year effects and the random group effects to zero.

\paragraph{Quad-based model}\label{quad-based-model}

Our quad-based model (QBM) perfectly mirrors its statistical description
(Eq. \#). We use the same approach for drawing parameter values as
described for the IPM.

@@

Adler, P. B., H. J. Dalgleish, and S. P. Ellner. 2012. Forecasting plant
community impacts of climate variability and change: when do competitive
interactions matter? Journal of Ecology 100:478--487.

@@

Chu, C., and P. B. Adler. 2015. Large niche differences emerge at the
recruitment stage to stabilize grassland coexistence. Ecological
Monographs.

@@

Rees, M., and S. P. Ellner. 2009. Integral projection models for
populations in temporally varying environments. Ecological Monographs
79:575--594.
